---
title: "TidyTuesday 2026 Week 13 — Ocean Temperature Explorer"
subtitle: "Birchy Head, Nova Scotia · Feb 2018 – Dec 2025"
author: "Siddhesh Pujari"
date: "2026-03-31"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## The Story

Off the coast of Nova Scotia, a network of underwater sensors has been quietly
recording ocean temperatures at **Birchy Head** since February 2018. Sensors sit
at seven fixed depths — from 2 m down to 40 m at low tide — logging readings
every day for nearly eight years.

This analysis explores two questions:

1. **How have temperatures changed over time?** — seasonal patterns, year-to-year
   variation, and whether any trends look different at different depths.
2. **How does temperature change with depth?** — the vertical structure of the
   water column and how it shifts across seasons.

---

## 1. Load Data

```{r packages, message=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(tidyr)
library(forcats)
```

```{r load-data}
tuesdata <- tidytuesdayR::tt_load('2026-03-31')

ocean_temperature             <- tuesdata$ocean_temperature
ocean_temperature_deployments <- tuesdata$ocean_temperature_deployments
```

Two datasets are used throughout:

- **`ocean_temperature`** — 19,165 rows of daily mean temperature (°C), standard
  deviation, and observation count, for each date × depth combination.
- **`ocean_temperature_deployments`** — 14 rows describing when and where each
  sensor package was deployed and retrieved.

---

## 2. Is the Change Seasonal? (Month × Year Heatmap)

The simplest way to look at eight years of data at once is a calendar heatmap —
each cell is the mean temperature for a given month and year, averaged across all
depths. This immediately shows whether the seasonal rhythm (cold winters, warm
summers) is consistent, and whether some years run warmer or cooler than others.

```{r seasonal-heatmap}
ocean_temperature |>
  mutate(
    year  = as.factor(year(date)),
    month = as.factor(month(date, label = TRUE))
  ) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(year, month)
  ) |>
  ggplot(aes(x = year, y = month, fill = mean_temp)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    title    = "Monthly Mean Temperature by Year",
    subtitle = "Averaged across all sensor depths",
    fill     = "°C",
    x        = "Year",
    y        = "Month"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank())
```

The seasonal pattern is strong and consistent every year: temperatures peak in
late summer (August–September) and bottom out in late winter (February–March).
Some years appear slightly warmer or cooler overall — 2021–2023 look warmer than
the surrounding years — but this is an association in the record, not a trend we
can attribute to any cause.

---

## 3. Seasonal Temperature Cycles by Depth

Does the seasonal pattern look the same at every depth, or does depth dampen or
delay the signal? Here we plot the monthly mean temperature separately for each of
the seven sensor depths.

```{r seasonal-by-depth}
ocean_temperature |>
  mutate(month = as.factor(month(date, label = TRUE))) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(month, sensor_depth_at_low_tide_m)
  ) |>
  ggplot(aes(
    x     = month,
    y     = mean_temp,
    color = factor(sensor_depth_at_low_tide_m),
    group = sensor_depth_at_low_tide_m
  )) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(
    title    = "Monthly Mean Temperature by Depth",
    subtitle = "Each line is one sensor depth — all years combined",
    color    = "Depth (m)",
    y        = "Mean Temp (°C)",
    x        = NULL
  ) +
  theme_minimal(base_size = 13)
```

Two things stand out:

- **Shallower sensors run warmer in summer.** The gap between 2 m and 40 m is
  largest in the warmest months, with the shallowest sensor peaking roughly
  3–4 °C above the deepest.
- **In winter the lines converge.** Below around 4 °C all depths track each other
  closely, suggesting the water column is well-mixed when temperatures are lowest.

---

## 4. Year-over-Year Variation by Depth

Zooming out to annual means lets us look for longer-term patterns. Each line here
tracks one depth level across the eight years of the record.

```{r annual-trend}
ocean_temperature |>
  mutate(year = year(date)) |>
  summarise(
    annual_mean = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  ggplot(aes(
    x     = year,
    y     = annual_mean,
    color = factor(sensor_depth_at_low_tide_m)
  )) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title    = "Annual Mean Temperature by Depth",
    subtitle = "One point per depth per year",
    color    = "Depth (m)",
    y        = "Mean Temp (°C)",
    x        = NULL
  ) +
  theme_minimal(base_size = 13)
```

The lines move largely in sync — years that were warmer at the surface also tend
to be warmer at depth. The 40 m sensor is a notable exception at times, showing
more independent variation than the shallower depths. Remember that some years
have gaps in certain depth records (we'll look at this in Section 6), so
year-to-year dips may partly reflect missing data rather than true cooling.

---

## 5. Annual Temperature Anomaly by Depth

Rather than looking at raw temperatures, anomalies show how each year compares to
the *long-term average for that month at that depth*. Calculating anomalies this
way removes the seasonal cycle, leaving behind only the year-to-year departures.

```{r anomaly-heatmap}
ocean_temperature |>
  mutate(month = month(date), year = year(date)) |>
  mutate(
    baseline = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(sensor_depth_at_low_tide_m, month)
  ) |>
  mutate(anomaly = mean_temperature_degree_c - baseline) |>
  summarise(
    mean_anomaly = mean(anomaly, na.rm = TRUE),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  mutate(depth_label = paste0(sensor_depth_at_low_tide_m, " m")) |>
  ggplot(aes(
    x    = factor(year),
    y    = reorder(depth_label, -sensor_depth_at_low_tide_m),
    fill = mean_anomaly
  )) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "white",
    high     = "#d6604d",
    midpoint = 0,
    name     = "Anomaly (°C)"
  ) +
  labs(
    title    = "Annual Temperature Anomaly by Depth",
    subtitle = "Deviation from each depth's long-term monthly mean",
    x        = "Year",
    y        = "Sensor Depth"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank())
```

Key patterns in the anomaly record:

- **2021–2023 are consistently red.** Warmer-than-baseline readings appear at
  nearly every depth in these years, which is consistent with the heatmap in
  Section 2 — though what drove this association is beyond the scope of this
  dataset.
- **2019 is consistently blue.** A cooler-than-baseline signal runs across all
  depths, again a station-wide pattern.
- **40 m in 2022** shows its strongest warm anomaly — darker red than the
  shallower depths that year, suggesting the deepest sensor sometimes behaves
  differently from those above it.
- **2024 — cooling at depth, near-neutral at surface.** The 20–40 m sensors flip
  negative while the 2–5 m sensors stay close to zero. *This signal coincides with
  significant data gaps in the depl_12 deployment (Section 7) and should be
  interpreted with caution.*
- **2018 has blank cells at shallow depths** — the record was incomplete in the
  first year of deployment.

---

## 6. How Does Temperature Change with Depth?

This plot shows the vertical temperature profile for each of the four seasons,
combining all years. Reading each line from bottom (40 m) to top (2 m) gives a
snapshot of how temperature is distributed through the water column.

```{r depth-profile}
ocean_temperature |>
  mutate(season = case_when(
    month(date) %in% c(12, 1, 2)  ~ "Winter",
    month(date) %in% c(3, 4, 5)   ~ "Spring",
    month(date) %in% c(6, 7, 8)   ~ "Summer",
    month(date) %in% c(9, 10, 11) ~ "Fall"
  )) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(season, sensor_depth_at_low_tide_m)
  ) |>
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) |>
  ggplot(aes(
    x     = mean_temp,
    y     = factor(sensor_depth_at_low_tide_m, levels = c(40, 30, 20, 15, 10, 5, 2)),
    color = season,
    group = season
  )) +
  geom_line(orientation = "y", linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c(
    "Winter" = "#2166ac",
    "Spring" = "#74add1",
    "Summer" = "#d6604d",
    "Fall"   = "#f4a582"
  )) +
  labs(
    title    = "Temperature Profile by Depth & Season",
    subtitle = "All years combined — lines read from deepest (bottom) to shallowest (top)",
    x        = "Mean Temp (°C)",
    y        = "Depth (m)",
    color    = NULL
  ) +
  theme_minimal(base_size = 13)
```

Three clear patterns emerge:

- **Summer & Fall: steep gradient.** The surface is 6–7 °C warmer than the
  bottom, with a well-defined thermocline in the upper water column.
- **Winter & Spring: nearly vertical lines.** All depths converge around 3–4 °C,
  meaning the water column is thoroughly mixed and little gradient remains.
- **Fall is the most stratified season** — the surface retains summer warmth while
  deep water has already cooled, producing the steepest gradient of all four
  seasons.

---

## 7. Data Quality — Where Are the Gaps?

Before drawing any conclusions it is worth knowing how complete the record is.
The TidyTuesday readme flags that some days have no data; here we build the full
expected grid of date × depth combinations and find what is missing.

```{r missing-grid}
full_grid <- tidyr::expand_grid(
  date                       = seq(min(ocean_temperature$date),
                                   max(ocean_temperature$date),
                                   by = "day"),
  sensor_depth_at_low_tide_m = c(2, 5, 10, 15, 20, 30, 40)
)

missing <- full_grid |>
  dplyr::anti_join(
    ocean_temperature,
    by = c("date", "sensor_depth_at_low_tide_m")
  ) |>
  dplyr::mutate(year = lubridate::year(date))
```

### 7a. Missing depth-days per year

```{r missing-summary}
missing |>
  dplyr::summarise(
    missing_depth_days = n(),
    missing_dates      = n_distinct(date),
    .by = year
  ) |>
  dplyr::arrange(year)
```

### 7b. Which depths are affected?

```{r missing-by-depth}
missing |>
  dplyr::summarise(
    missing_days = n(),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  tidyr::pivot_wider(
    names_from   = sensor_depth_at_low_tide_m,
    values_from  = missing_days,
    values_fill  = 0,
    names_prefix = "d"
  ) |>
  dplyr::arrange(year)
```

Only the **2 m, 5 m, and 40 m** sensors have gaps — the other four depths have
complete records across the full eight years. This rules out full station outages;
individual sensors are the issue.

### 7c. Missing data heatmap

This heatmap shows the percentage of days missing in each month, for each depth.
White = fully complete; deep red = mostly missing.

```{r missing-heatmap}
full_grid |>
  dplyr::left_join(
    ocean_temperature |>
      dplyr::select(date, sensor_depth_at_low_tide_m) |>
      dplyr::mutate(has_data = TRUE),
    by = c("date", "sensor_depth_at_low_tide_m")
  ) |>
  dplyr::mutate(
    has_data = tidyr::replace_na(has_data, FALSE),
    month    = lubridate::floor_date(date, "month")
  ) |>
  dplyr::summarise(
    pct_missing = mean(!has_data) * 100,
    .by = c(month, sensor_depth_at_low_tide_m)
  ) |>
  ggplot(aes(
    x    = month,
    y    = factor(sensor_depth_at_low_tide_m),
    fill = pct_missing
  )) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low  = "white",
    high = "#c0392b",
    name = "% days\nmissing"
  ) +
  labs(
    title    = "Missing Data by Depth & Month",
    subtitle = "Only 2 m, 5 m, and 40 m sensors show gaps",
    x        = NULL,
    y        = "Depth (m)"
  ) +
  theme_minimal(base_size = 13)
```

### 7d. Linking gaps to specific deployments

```{r missing-deployments}
missing |>
  summarise(
    gap_start = min(date),
    gap_end   = max(date),
    days      = n(),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  arrange(year, sensor_depth_at_low_tide_m) |>
  left_join(
    ocean_temperature_deployments |>
      select(deployment_id, start_date, end_date),
    by = join_by(between(gap_start, start_date, end_date))
  )
```

Every gap maps to a named deployment, confirming these are not random sensor
failures but specific sensors that were absent during active deployments:

- **depl_02 (2019)** — the 2 m sensor dropped out 14 days before retrieval and
  was not reinstalled until the following November (146 days total).
- **depl_09 (2022–2023)** — the 40 m sensor was absent for almost the entire
  deployment: 192 days across two calendar years.
- **depl_10 (2023)** — the 5 m sensor was missing for 131 of the 179 deployment
  days.
- **depl_12 (2024)** — the worst deployment: three sensors affected simultaneously.
  The 2 m sensor was absent for 202 days; the 40 m sensor dropped out after just
  33 days; the 5 m sensor fell silent near the end (51 days). **The 2024
  cooling anomaly at depth seen in Section 5 coincides directly with these gaps —
  that signal warrants extra caution.**
- **depl_14 (2025)** — the 5 m sensor drops out nine days before the end of the
  record.

A better way to visualise above is through the below chart:

```{r}
#| echo: false
# Build the full expected grid and find missing date × depth combinations
full_grid <- tidyr::expand_grid(
  date                       = seq(min(ocean_temperature$date),
                                   max(ocean_temperature$date),
                                   by = "day"),
  sensor_depth_at_low_tide_m = c(2, 5, 10, 15, 20, 30, 40)
)

missing <- full_grid |>
  dplyr::anti_join(ocean_temperature,
                   by = c("date", "sensor_depth_at_low_tide_m")) |>
  dplyr::mutate(year = lubridate::year(date))

# Summarise gaps into continuous start–end segments per deployment × depth
gap_segments <- missing |>
  dplyr::left_join(
    ocean_temperature_deployments |>
      dplyr::select(deployment_id, start_date, end_date),
    by = dplyr::join_by(between(date, start_date, end_date))
  ) |>
  dplyr::summarise(
    gap_start = min(date),
    gap_end   = max(date),
    .by = c(deployment_id, sensor_depth_at_low_tide_m)
  ) |>
  dplyr::mutate(depth_label = paste0(sensor_depth_at_low_tide_m, " m"))

# Deployment bars — sorted chronologically for the y-axis
deployments_ordered <- ocean_temperature_deployments |>
  dplyr::arrange(start_date) |>
  dplyr::mutate(deployment_id = forcats::fct_inorder(deployment_id))

# Match factor levels in gap_segments
gap_segments <- gap_segments |>
  dplyr::mutate(deployment_id = factor(deployment_id,
                                       levels = levels(deployments_ordered$deployment_id)))

gantt <- ggplot() +
  # Grey background bars = full deployment span
  geom_segment(
    data = deployments_ordered,
    aes(x = start_date, xend = end_date,
        y = deployment_id, yend = deployment_id),
    colour = "grey80", linewidth = 4
  ) +
  # Coloured segments = data gaps, by depth
  geom_segment(
    data = gap_segments,
    aes(x = gap_start, xend = gap_end,
        y = deployment_id, yend = deployment_id,
        colour = depth_label),
    linewidth = 4, alpha = 0.85
  ) +
  scale_colour_manual(
    values = c("2 m" = "#e41a1c", "5 m" = "#ff7f00", "40 m" = "#377eb8"),
    name   = "Gap depth"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title    = "Sensor Deployment Timeline & Data Gaps",
    subtitle = "Grey = active deployment  |  Coloured = missing data by depth",
    x        = NULL,
    y        = "Deployment ID"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

gantt
```

```{r}
#| echo: false
#| include: false

# ggsave("deployment_gantt.png", plot = gantt,
      #  width = 10, height = 5, dpi = 150, bg = "white")

# ggsave("month_x_year.png", plot = month_year,
# width = 10, height = 5, dpi = 150, bg = "white")
```
Each row is one sensor deployment, ordered top-to-bottom from earliest to latest. The grey bar spans the full deployment period. Coloured segments mark stretches where data was missing for a specific depth — red for 2 m, orange for 5 m, blue for 40 m.

At a glance you can see that gaps are rare and isolated to three depths only, with depl_12 (2024) being the worst-affected — three depths failing simultaneously, which is why the 2024 anomaly signal in Section 5 needs to be treated with caution.


---

*Data source: Centre for Marine Applied Research (CMAR) Coastal Monitoring Program,
via [TidyTuesday 2026 Week 13](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-03-31).*
