---
title: "TidyTuesday 2026 Week 14 — Ocean Temperature (Birchy Head, Nova Scotia)"
author: ""
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This analysis uses the [TidyTuesday 2026 Week 14](https://github.com/rfordatascience/tidytuesday) dataset, which contains ocean temperature readings from a fixed monitoring station at
**Birchy Head, Nova Scotia**. Sensors are deployed at multiple depths (2 m to 40 m at
low tide), recording sub-daily temperature measurements from February 2018 to late 2025.

Two datasets are used:

- `ocean_temperature`: Daily mean/SD temperature by depth
- `ocean_temperature_deployments`: Metadata on each sensor deployment period

---

## 1. Load Packages

```{r packages}
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(tidyr)
```

---

## 2. Load Data from Tidytuesday 31st March 2026
```{r}
tuesdata <- tidytuesdayR::tt_load('2026-03-31')

ocean_temperature <- tuesdata$ocean_temperature
ocean_temperature_deployments <- tuesdata$ocean_temperature_deployments
```

```{r preview-deployments}
#| echo: false
#| include: false
# Sensor deployment metadata: 14 deployments
ocean_temperature_deployments
```
---

```{r}
#| echo: FALSE
#| include: FALSE
ocean_temperature |>
  mutate(year = year(date)) |>
  summarise(
    annual_mean = mean(mean_temperature_degree_c, na.rm = TRUE),
    annual_sd   = sd(mean_temperature_degree_c,   na.rm = TRUE),
    .by = year
  ) 
```

## Is the change seasonal ? (month x year heatmap)
```{r}
ocean_temperature |>
  mutate(year  = as.factor(year(date)),
         month = as.factor(month(date, label = TRUE))
         ) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(year, month)
  ) |>
  ggplot(aes(x = year, y = month, fill = mean_temp)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Monthly Mean Temperature by Year",
       fill  = "°C", x = "Year", y = "Month")
```


```{r}
#| echo: FALSE
#| include: FALSE
ocean_temperature |>
  mutate(year  = as.factor(year(date)),
         month = as.factor(month(date, label = TRUE)),
         ) |>
  mutate(season = case_when(
    month(date) %in% c(12, 1, 2) ~ "Winter",
    month(date) %in% c(3, 4, 5)  ~ "Spring",
    month(date) %in% c(6, 7, 8)  ~ "Summer",
    month(date) %in% c(9, 10, 11) ~ "Fall"),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
    ) |> 
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(year, season)
  ) |>
  ggplot(aes(x = year, y = season, fill = mean_temp)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Seasonal Mean Temperature by Year",
       fill  = "°C", x = "Year", y = "Season")
```

```{r}
#| echo: FALSE
#| include: FALSE
# Reversed for ggplot (which draws factor levels bottom-to-top)
# We want Dec at top → Nov at bottom, so levels must be Nov...Dec
month_order <- rev(c("Dec", "Jan", "Feb", "Mar", "Apr", "May",
                     "Jun", "Jul", "Aug", "Sep", "Oct", "Nov"))
# levels: Nov=1(bottom), Oct=2, ..., Jan=11, Dec=12(top)

ocean_temperature |>
  mutate(
    year  = as.factor(year(date)),
    month = month(date, label = TRUE, abbr = TRUE),
    month = factor(month, levels = month_order)
  ) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(year, month)
  ) |>
  ggplot(aes(x = year, y = month, fill = mean_temp)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", name = "°C") +
  # y positions: Nov=1, Oct=2, Sep=3, Aug=4, Jul=5, Jun=6,
  #              May=7, Apr=8, Mar=9, Feb=10, Jan=11, Dec=12
  # Winter: Dec(12), Jan(11), Feb(10)
  annotate("segment", x = 0.4, xend = 0.4, y = 9.6, yend = 12.4,
           colour = "#0C7BDC", linewidth = 1.2,
           arrow = arrow(ends = "both", length = unit(4, "pt"), type = "closed")) +
  annotate("text", x = 0.25, y = 11, label = "Winter",
           colour = "#0C7BDC", size = 3, hjust = 1, fontface = "italic") +
  # Spring: Mar(9), Apr(8), May(7)
  annotate("segment", x = 0.4, xend = 0.4, y = 6.6, yend = 9.4,
           colour = "#009E73", linewidth = 1.2,
           arrow = arrow(ends = "both", length = unit(4, "pt"), type = "closed")) +
  annotate("text", x = 0.25, y = 8, label = "Spring",
           colour = "#009E73", size = 3, hjust = 1, fontface = "italic") +
  # Summer: Jun(6), Jul(5), Aug(4)
  annotate("segment", x = 0.4, xend = 0.4, y = 3.6, yend = 6.4,
           colour = "#E66100", linewidth = 1.2,
           arrow = arrow(ends = "both", length = unit(4, "pt"), type = "closed")) +
  annotate("text", x = 0.25, y = 5, label = "Summer",
           colour = "#E66100", size = 3, hjust = 1, fontface = "italic") +
  # Fall: Sep(3), Oct(2), Nov(1)
  annotate("segment", x = 0.4, xend = 0.4, y = 0.6, yend = 3.4,
           colour = "#5D3A9B", linewidth = 1.2,
           arrow = arrow(ends = "both", length = unit(4, "pt"), type = "closed")) +
  annotate("text", x = 0.25, y = 2, label = "Fall",
           colour = "#5D3A9B", size = 3, hjust = 1, fontface = "italic") +
  coord_cartesian(clip = "off") +
  labs(
    title = "Monthly Mean Temperature by Year",
    x = "Year", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.margin  = margin(5, 10, 5, 60),
    axis.text.x  = element_text(angle = 45, hjust = 1),
    panel.grid   = element_blank()
  )


```

## 4. Seasonal Temperature Cycles by Depth

Monthly mean temperature grouped by sensor depth, showing the seasonal warming and
cooling pattern across all depth levels.

```{r seasonal-by-depth}
ocean_temperature |>
  mutate(month = as.factor(month(date, label = TRUE))) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c),
    .by = c(month, sensor_depth_at_low_tide_m)
  ) |>
  ggplot(aes(
    x     = month,
    y     = mean_temp,
    color = factor(sensor_depth_at_low_tide_m),
    group = sensor_depth_at_low_tide_m
  )) +
  geom_line() +
  labs(
    title = "Monthly Mean Temperature by Depth",
    color = "Depth (m)",
    y     = "Mean Temp (°C)",
    x     = NULL
  )
```

1. As expected, at the lower depths mean temp is at the higher levels and it goes down as depth increases
2. During the same months when high temps are expected, at lower depths mean temp is high

---

## 5. Year-over-Year Warming Trend

Annual mean temperature per depth level — useful for detecting long-term warming or
cooling signals at different depths.

```{r annual-trend}
ocean_temperature |>
  mutate(year = year(date)) |>
  summarise(
    annual_mean = mean(mean_temperature_degree_c),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  ggplot(aes(
    x     = year,
    y     = annual_mean,
    color = factor(sensor_depth_at_low_tide_m)
  )) +
  geom_line() +
  geom_point() +
  labs(
    title = "Annual Mean Temperature by Depth",
    color = "Depth (m)",
    y     = "Mean Temp (°C)",
    x     = NULL
  )
```

```{r}

ocean_temperature |>
  mutate(month = month(date), year = year(date)) |>
  mutate(
    baseline = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(sensor_depth_at_low_tide_m, month)
  ) |>
  mutate(anomaly = mean_temperature_degree_c - baseline) |>
  summarise(
    mean_anomaly = mean(anomaly, na.rm = TRUE),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  mutate(depth_label = paste0(sensor_depth_at_low_tide_m, " m")) |>
  ggplot(aes(x = factor(year), y = reorder(depth_label, -sensor_depth_at_low_tide_m), fill = mean_anomaly)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "white",
    high     = "#d6604d",
    midpoint = 0,
    name     = "Anomaly (°C)"
  ) +
  labs(
    title    = "Annual Temperature Anomaly by Depth",
    subtitle = "Deviation from each depth's long-term monthly mean",
    x        = "Year",
    y        = "Sensor Depth"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank())

```
---

Here's what the heatmap reveals:

Key Patterns
Coherent warm/cool years across all depths
2021–2023 are consistently red (warmer than baseline) and 2019 is consistently blue (cooler) — this signal runs through all depths simultaneously, suggesting large-scale oceanographic or climate forcing rather than surface heating alone.

40 m stands out in 2022
The deepest sensor has its strongest warm anomaly in 2022, darker red than the shallower depths that year. This confirms the earlier statistical finding — the 40 m layer has its own variability that doesn't always track the surface.

2024 — cooling at depth, near-neutral at surface
The bottom layers (20–40 m) flip strongly negative in 2024 while the surface (2–5 m) stays near zero. This is a clear depth-dependent signal — whatever drove the 2024 cooling affected deep water more than the surface.

2018 is missing data at some depths
The blank cells at shallow depths in 2018 reflect incomplete deployment coverage early in the record — worth keeping in mind when interpreting that year.

## 6. Thermal Stratification (Surface vs. Deep)

Difference between 2 m and 40 m daily temperatures. Positive values indicate a warmer
surface layer — a sign of thermal stratification, typically strongest in summer.

```{r stratification}
ocean_temperature |>
  filter(sensor_depth_at_low_tide_m %in% c(2, 40)) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c),
    .by = c(date, sensor_depth_at_low_tide_m)
  ) |>
  pivot_wider(
    names_from   = sensor_depth_at_low_tide_m,
    values_from  = mean_temp,
    names_prefix = "depth_"
  ) |>
  mutate(stratification = depth_2 - depth_40) |>
  ggplot(aes(x = date, y = stratification)) +
  geom_line(alpha = 0.5) +
  geom_smooth() +
  labs(
    title = "Thermal Stratification (2 m − 40 m)",
    y     = "Temp Difference (°C)",
    x     = NULL
  )
```

---

## Q2 : How does temperature change with depth

```{r}
ocean_temperature |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    median_temp = median(mean_temperature_degree_c, na.rm = TRUE),
    sd_temp = sd(mean_temperature_degree_c, na.rm = TRUE),
    min_temp = min(mean_temperature_degree_c, na.rm = TRUE),
    max_temp = max(mean_temperature_degree_c, na.rm = TRUE),
    .by = sensor_depth_at_low_tide_m
  ) |>
  arrange(sensor_depth_at_low_tide_m)

```

```{r}
ocean_temperature |>
  mutate(season = case_when(
    month(date) %in% c(12, 1, 2) ~ "Winter",
    month(date) %in% c(3, 4, 5)  ~ "Spring",
    month(date) %in% c(6, 7, 8)  ~ "Summer",
    month(date) %in% c(9, 10, 11) ~ "Fall"
  )) |>
  summarise(
    mean_temp = mean(mean_temperature_degree_c, na.rm = TRUE),
    .by = c(season, sensor_depth_at_low_tide_m)
  ) |>
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) |>
  # ggplot(aes(x = mean_temp, y = factor(sensor_depth_at_low_tide_m))) +
  ggplot(aes(x = mean_temp, y = factor(sensor_depth_at_low_tide_m, levels = c(40, 30, 20, 15, 10, 5, 2 )))) +
  geom_line(aes(group = season), orientation = "y") +
  geom_point(aes(color = season), size = 3) +
  scale_color_manual(values = c("Winter" = "#2166ac", "Spring" = "#74add1",
                                 "Summer" = "#d6604d", "Fall"   = "#f4a582")) +
  labs(title   = "Temperature Profile by Depth & Season",
       x       = "Mean Temp (°C)",
       y       = "Depth (m)",
       color   = NULL) +
  theme_minimal(base_size = 13)

```

Three clear takeaways:

Summer & Fall: Strong gradient — surface is much warmer than the bottom (~6–7°C difference). The thermocline is well-established
Winter & Spring: Nearly vertical lines — all depths converge around 3–4°C, meaning the water column is fully mixed with almost no gradient
Fall is the most stratified season — surface retains summer heat while deep water has already cooled, creating the steepest gradient of all four seasons


## 7. Deployment Duration Analysis

Bar chart of how long each sensor deployment lasted, ordered by start date.

```{r deployment-duration}
ocean_temperature_deployments |>
  mutate(duration_days = as.numeric(end_date - start_date)) |>
  ggplot(aes(x = start_date, y = duration_days)) +
  geom_col() +
  labs(
    title = "Sensor Deployment Durations",
    x     = "Start Date",
    y     = "Days"
  )
```

---

## 8. High-Variability Days (Anomaly Hunting)

Days where within-day temperature SD is in the top 1% — potential anomalies, mixing
events, or sensor issues worth investigating.

```{r anomaly-days}
ocean_temperature |>
  filter(sd_temperature_degree_c >
    quantile(sd_temperature_degree_c, 0.99, na.rm = TRUE)) |>
  arrange(desc(sd_temperature_degree_c))
```

## 9. Missing Data Investigation

The TidyTuesday readme warns to be wary of days with no data. We investigate
whether gaps are random or systematic.

### 9a. Build the full expected grid and find missing combinations

```{r missing-grid}
# Every date × depth combination we expect to see
full_grid <- tidyr::expand_grid(
  date                       = seq(min(ocean_temperature$date),
                                   max(ocean_temperature$date), by = "day"),
  sensor_depth_at_low_tide_m = c(2, 5, 10, 15, 20, 30, 40)
)

# Date and depths present in full_grid but absent from ocean_temperature data
missing <- full_grid |>
  dplyr::anti_join(ocean_temperature,
                   by = c("date", "sensor_depth_at_low_tide_m")) |>
  dplyr::mutate(year = lubridate::year(date))
```

### 9b. Summary — missing depth-days per year

```{r missing-summary}
missing |>
  dplyr::summarise(
    missing_depth_days = n(),
    missing_dates      = n_distinct(date),
    .by = year
  ) |>
  dplyr::arrange(year)
```

### 9c. Which depths are affected?

```{r missing-by-depth}
# Pivot wider to see which depth columns have gaps per year
missing |>
  dplyr::summarise(
    missing_days = n(),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  tidyr::pivot_wider(
    names_from   = sensor_depth_at_low_tide_m,
    values_from  = missing_days,
    values_fill  = 0,
    names_prefix = "d"
  ) |>
  dplyr::arrange(year)
```

Only depths 2 m, 5 m, and 40 m have gaps — the other four depths have complete
records. This rules out full station outages; **individual sensors are the problem**.

### 9d. Heatmap — missing data by depth and month

```{r missing-heatmap}
full_grid |>
  dplyr::left_join(
    ocean_temperature |>
      dplyr::select(date, sensor_depth_at_low_tide_m) |>
      dplyr::mutate(has_data = TRUE),
    by = c("date", "sensor_depth_at_low_tide_m")
  ) |>
  dplyr::mutate(
    has_data = tidyr::replace_na(has_data, FALSE),
    month    = lubridate::floor_date(date, "month")
  ) |>
  dplyr::summarise(
    pct_missing = mean(!has_data) * 100,
    .by = c(month, sensor_depth_at_low_tide_m)
  ) |>
  ggplot(aes(x = month,
             y = factor(sensor_depth_at_low_tide_m),
             fill = pct_missing)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#c0392b",
                      name = "% days\nmissing") +
  labs(title = "Missing Data by Depth & Month",
       x     = NULL,
       y     = "Depth (m)") +
  theme_minimal()
```

### 9e. Cross-check gaps against deployment periods

```{r missing-deployments}
# Join each gap to the deployment it falls within
missing |>
  summarise(
    gap_start = min(date),
    gap_end   = max(date),
    days      = n(),
    .by = c(year, sensor_depth_at_low_tide_m)
  ) |>
  arrange(year, sensor_depth_at_low_tide_m) |>
  left_join(
    ocean_temperature_deployments |>
      select(deployment_id, start_date, end_date),
    by = join_by(between(gap_start, start_date, end_date))
  )
```

Every single gap maps exactly to a named deployment period — confirming these
are **not sensor failures but missing sensors during active deployments**.

Key findings:

- **depl_02 (2019)** — 2 m sensor dropped out on 2019-04-18, fourteen days
  *before* retrieval on 2019-05-02, and was not included in the next deployment
  until 2019-11-21 (146 days total, straddling the end of depl_02 and the start
  of the following period)
- **depl_09 (2022–2023)** — 40 m sensor was absent for almost the entire
  deployment (2022-11-03 to 2023-05-15): gap runs 2022-11-04 to 2023-05-14,
  192 days across two calendar years
- **depl_10 (2023)** — 5 m sensor missing for most of the deployment
  (2023-05-15 to 2023-11-10): gap runs 2023-07-02 to 2023-11-09, covering
  131 of the 179 deployment days
- **depl_12 (2024)** — worst deployment, three sensors affected simultaneously:
  - 2 m absent for virtually the entire deployment (2024-05-03 to 2024-11-20, 202 days)
  - 40 m present only at the very start (2024-05-03 to 2024-06-04, then gone for 33 days)
  - 5 m drops out near the end (2024-10-01 to 2024-11-20, 51 days)
- **depl_14 (2025)** — 5 m sensor drops out nine days before the end of the
  record (2025-11-28 to 2025-12-06)

The **2024 cooling anomaly at depth** seen in the annual anomaly heatmap
(Section 5) coincides with the depl_12 gaps — that signal should be interpreted
with caution.
