---
title: "Exploring Repair cafes"
subtitle: "From tidytuesday data"
format: html
---

# Loading libraries and data
```{r loading lib and data}
#| code-fold: true
#| output: false
tuesdata <- tidytuesdayR::tt_load('2026-04-07')

# load libraries
library(tidyverse)
library(conflicted)
library(admiral)
library(polyglotr)
library(countrycode)
library(rnaturalearth)
library(rnaturalearthdata)
library(plotly)
library(arrow) # for saving as parquet data
library(ggplot2)

# conflicts
conflicts_prefer(dplyr::filter)

repairs <- as_tibble(tuesdata$repairs) |> convert_blanks_to_na()
repairs_text <- as_tibble(tuesdata$repairs_text) |> convert_blanks_to_na()
```

# Prepare repairs data for further analysis
```{r prep repairs}
repairs <- repairs |>
  mutate(
    repair_year = as.factor(year(repair_date)),
    repair_month = floor_date(repair_date, "month"),
    repaired = if_else(repaired == "ja", "yes", repaired),
    # Get country , continent , country flags with countrycode package
    country_name = countrycode(
      country,
      origin = "iso2c",
      destination = "country.name"
    ),
    country_flag = countrycode(
      country,
      origin = "iso2c",
      destination = "unicode.symbol"
    ),
    continent = countrycode(
      country,
      origin = "iso2c",
      destination = "continent"
    )
  ) |>
  select(repair_id, repair_date, repair_year, everything())
```

# Prepare repairs text data

```{r prep repairs text}
# Update N/A to NA values to be analysed correctly
repairs_text <- repairs_text |>
  mutate(model = if_else(model == "N/A", NA_character_, model))
```

# No of cafes growth per year

```{r}
#| eval: false
#No of cafes growth each year ------------------------------------------------------
repairs |>
  group_by(repair_year) |>
  summarise(N_cafes = n_distinct(repair_cafe_name)) |>
  # mutate(repair_year = as.factor(repair_year)) |>
  ggplot(
    mapping = aes(x = repair_year, y = N_cafes)
  ) +
  geom_bar(stat = "identity")
```

# Map of all cafes -----
```{r }
#| echo: false
#| eval: false

# Create a map to mark where all the cafes are located
# --- Step 1: Count cafes per country from your data ---
cafe_counts <- repairs |>
  group_by(country_name) |>
  summarise(num_cafes = n_distinct(repair_cafe_name), .groups = "drop") |>
  arrange(desc(num_cafes))

cafe_counts <- cafe_counts |>
  mutate(
    country_name = case_when(
      country_name == "United States" ~ "United States of America",
      country_name == "French Guiana" ~ "French Guiana", # part of France in rnaturalearth, may stay unmatched
      country_name == "Hong Kong SAR China" ~ "Hong Kong",
      TRUE ~ country_name
    )
  )

# --- Step 2: Get world map polygons ---
world <- ne_countries(scale = "medium", returnclass = "sf")

# --- Step 3: Join your cafe counts to the map data ---
world_cafes <- world |>
  left_join(cafe_counts, by = c("name" = "country_name"))

# --- Step 4: Plot ---
ggplot(world_cafes) +
  geom_sf(aes(fill = num_cafes), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey85", # countries with no cafes appear grey
    name = "Number of Cafes"
  ) +
  labs(
    title = "Cafe Locations Around the World",
    subtitle = "By country"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 8),
    legend.position = "bottom"
  )
```

# Exploring categories and their repair status over the years

```{r}
#| echo: false
repaired_count_long <- repairs |>
  # drop any rows with missing category
  filter_out(is.na(category) | is.na(repaired)) |>
  group_by(repair_year, category, repaired) |>
  summarise(repaired_status = n(), .groups = "drop") |>
  mutate(repaired = factor(repaired, levels = c("yes", "half", "no")))

repaired_count_wide <- repaired_count_long |>
  pivot_wider(
    id_cols = c(repair_year, category),
    names_from = repaired,
    values_from = repaired_status
  )

electric_non_electric <- repaired_count_long |>
  filter(grepl("electric", category))
```

Many categories make it difficult to visualise all at once
For now, only visualize categories that are either electric and non electric

```{r}
#| echo: false
#| eval: false
plot_rep_st_category <- function(data, title) {
  p <- ggplot(
    data,
    aes(x = factor(repair_year), y = repaired_status, fill = repaired)
  ) +
    geom_col(position = "dodge") + # side-by-side bars, not stacked
    scale_fill_manual(
      values = c("yes" = "#2ecc71", "half" = "#f39c12", "no" = "#e74c3c"),
      name = "Repair Status"
    ) +
    facet_wrap(~category, ncol = 2) +
    labs(
      title = title,
      x = "Year",
      y = "Number of Products"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold", size = 8),
      axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
      legend.position = "bottom",
      panel.spacing = unit(0.8, "lines")
    )

  return(p)
}
```

Let's look at each category 
```{r}
#| echo: false
#| eval: false
plot_rep_st_household <- plot_rep_st_category(
  data = electric_non_electric |> filter(grepl("Household", category)),
  title = "Status for household appliances by electric/non over years"
)

plot_rep_st_household
```

```{r}
#| echo: false
#| eval: false
plot_rep_st_tools <- plot_rep_st_category(
  data = electric_non_electric |> filter(grepl("Tools", category)),
  title = "Status for tools by electric/non over years"
)

plot_rep_st_tools
```

```{r}
#| echo: false
#| eval: false

plot_rep_st_toys <- plot_rep_st_category(
  data = electric_non_electric |> filter(grepl("Toys", category)),
  title = "Status for toys by electric/non over years"
)

plot_rep_st_toys
```

# save all datasets
```{r}
#| include: false
#| eval: false

# Save all datasets as parquet
data_path <- here::here("2026_04_07_repair_cafes/data/")

write_parquet(repairs, paste0(data_path, "repairs.parquet"))
write_parquet(repairs_text, paste0(data_path, "repairs_text.parquet"))
# write_parquet(world_cafes, paste0(data_path, "world_cafes.parquet"))
saveRDS(world_cafes, paste0(data_path, "world_cafes.rds"))
write_parquet(electric_non_electric, paste0(data_path, "electric_non_electric.parquet"))

```

# Interactive shiny code

## Number of repairs
```{r}
# by country, repair cafe and year

repairs_by_month <- repairs |> 
  filter(country_name == "Germany") |> 
  mutate(repair_month = floor_date(repair_date, "month")) |> 
  count(repair_month, name = "no_of_repairs")

# Plot
p <- repairs_by_month |> 
  ggplot(aes(x = repair_month, y = no_of_repairs)) +
  geom_area(fill = "#e8734a", alpha = 0.3) +               # filled area
  geom_line(color = "#e8734a", linewidth = 0.8) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "Number of repairs in Germany",
    x = "Year",
    y = "Number of repairs"
  ) +
  theme_classic()

ggplotly(p) |> 
  style(hoverinfo = "none", traces = 1) |> 
  plotly::layout(
    hovermode = "x unified"
  )
```

## Most presented products in Germany

```{r}
most_products_pres <- repairs |> 
  filter(country_name == "Germany") |> 
  count(kind_of_product, name = "no_products_presented") |> 
  arrange(desc(no_products_presented)) |> 
  head(10) |> 
  mutate(kind_of_product = fct_reorder(kind_of_product, no_products_presented)) 

p <- most_products_pres |> 
  ggplot(aes(x = kind_of_product, y = no_products_presented, fill = kind_of_product)) +
  geom_bar(stat = "identity") +
  guides(fill = "none") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Most presented products in Germany",
    x = "",
    y = ""
  ) +
  theme(
    axis.ticks = element_blank(),
    panel.background  = element_rect(fill = "white", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.border      = element_blank(),    # removes the bw border box
    panel.grid.major.y = element_blank()    # optional: removes horizontal grid lines too
)

ggplotly(p)


```

```{r}
repairs |> 
group_by(brand) |> 
        summarise(
            no_products_presented = n(),
            success_rate = sum(repaired == "yes") / n(),   # swap formula here
            .groups = "drop"
        ) |> 
          arrange(desc(no_products_presented)) |> 
            head(10) |> 
            mutate(brand = fct_reorder(brand, no_products_presented))
```
